Optimal updating factor in adaptive flat-distribution simulations

### 2016-04-26 doc/invt.tex
new xerr.gp, edits

Here we shall call a simulation that targets
a nearly flat distribution\cite{
dayal2004, trebst2004, singh2011},
$p_i$,
a flat-distribution sampling (FDS) simulation.
%
Roughly speaking, there are two types of FDS methods.
%
The first type,
referred to as the equilibrium FDS, % (E-FDS) methods hereinafter,
uses a fixed bias potential, $V_i$,
(which is estimated beforehand)
during simulation.
%
Examples include the original
entropic or multicanonical sampling\cite{
berg1992, lee1993}
and simulated tempering\cite{
marinari1992, lyubartsev1992}
with $i$ being the index of energy
and that of temperature, respectively.
%
The estimated bias potential, $V_i$,
are often improved
by the histogram, $H_i$, along $i$ accumulated
during simulation.
%
Using $H_i$ as an estimate of $\pi_i$ in
Eq. \eqref{eq:pi_p_phi1}
allows us to correct $V_i$ as
%
\begin{equation}
V^\mathrm{corr}_i
=
V_i
+
\ln \frac{ H_i }
         { p_i }.
\label{eq:vcorr_equil}
\end{equation}
%
\note{This follows from
  $$
  \begin{aligned}
    H_i \approx \pi_i
    &\propto p^*_i \, e^{-V_i},
    \\
    p_i
    &\propto p^*_i \, e^{-V^\mathrm{corr}_i}.
  \end{aligned}
  $$
}
For an unknown system,
it is often difficult
to guess a sufficiently accurate
$V_i$ to start with.
%
Thus,
it can take many iterative rounds of simulations
to achieve the desired flat distribution.
%
\note{Further, a reasonably precise histogram
requires an accumulation period
longer than the autocorrelation time,
which can also be hard to estimate beforehand.}

@@

In our test system,
$i$ represents a one-dimensional
non-periodic variable,
and $n = 100$.
%
Both the target and intrinsic distributions
are assumed to be flat:
$p^*_i = p_i = 1/n$.
%
%We wish to see how
%the schedule, $\alpha(t)$,
%affects the fluctuating error
%of the bias potential
%at the end of the simulation.
%
Initially,
we set the bias potential to zero,
$V_i \equiv 0$,
%
and equilibrate the system for $10^7$ steps
under a constant updating magnitude,
$\alpha_0 = 10^{-4}$,
unless specified otherwise.
%
After this,
we reset the origin of the time, $t$,
and start the testing schedule.
%specified by either
%Eq. \eqref{eq:q_opt}
%or
%Eq. \eqref{eq:alpha_invtlambda}.

@@

We performed adaptive FDS simulations
using the single-bin updating scheme
and the triple-bin updating scheme with $\mu_1 = 0.24$.
%
For each value of $\lambda$, we ran $1000$ independent simulations
of $10^8$ steps, and averaged the results.
%
As shown in Fig. \ref{fig:errinvt},
the errors from the simulations
agreed well those from the theory.

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=1.0\linewidth]{fig/errinvt.pdf}
  }
  \caption{
    \label{fig:errinvt}
    Error $E$ versus $\lambda$
    in Eq. \eqref{eq:alpha_invtlambda}
    for the inverse-time schedule.
    %
    The points are results from averaging over $1000$ independent runs;
    the curves are theoretical predictions.
    %
    \note{The figure was produced by \texttt{doc/fig/errinvt.gp}.
      The data were produced by the script \texttt{data/invtcscan.py},
      and saved under \texttt{data/invt}
      (see the \texttt{Makefile} there).
    }%
  }
\end{center}
\end{figure}

@@

We computed the optimal schedule from Eq. \eqref{eq:q_opt}
for the Gaussian (with $\sigma = 10$)
and bandpass (with $K = 4$) updating schemes.
%
As shown in Fig. \ref{fig:optacmp},
while the optimal schedules of the bandpass scheme
were always given by the inverse-time formula,
those of the Gaussian updating scheme was complex
and dependent on the sampling process.

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=0.95\linewidth]{fig/optacmp.pdf}
  }
  \caption{
    \label{fig:optacmp}
    Optimal schedules from Eq. \eqref{eq:q_opt},
    versus the optimized inverse-time schedule
    Eq. \eqref{eq:alpha_invtlambda}
    for the Gaussian updating scheme
    with $\sigma = 10$
    and the local updating process.
    %
    \note{This figure was produced by \texttt{doc/fig/optacmp.gp}.
      The data were computed by the program \texttt{prog/predict},
      see particularly the function \texttt{intq\_save()}
      in \texttt{prog/intq.h}.
      The data were saved under the directory \texttt{data/opta}.
      See the \texttt{make} object \texttt{sigsingle}
      in the \texttt{Makefile} there.
    }%
  }
\end{center}
\end{figure}

For the Gaussian updating scheme,
we listed the errors from different schedules,
in Table \ref{tab:error_Gaussian}.
%
Clearly,
the optimal schedules yielded the smallest errors.

@@

To verify Eq. \eqref{eq:half_alpha0},
we computed the optimal schedules
for different values of $q(T)$,
and plotted the final error
against the initial updating magnitude, $\alpha(0)$.
%
As shown in Fig. \ref{fig:erriascan},
The minimal error is reached at
$\alpha(0) = \alpha/2 = 5 \times 10^{-5}$.

@@

In contrast,
the normalized error of the bandpass updating schemes
were insensitive to the simulation length,
in agreement with Eq. \eqref{eq:error_sinc}.
%
Besides, the bandpass updating schemes
yielded smaller errors than
the corresponding Gaussian ones.

@@

Each adaptive FDS method is associated with
an updating scheme,
or mathematically an updating matrix.
%whose elements $w_{ij}$ specifies the
%relative magnitude of updating the bias potential
%at bin $i$ upon a visit to bin $j$.
%
Our method computes the optimal schedule from
the eigenvalues of the updating matrix,
and the integrals of the autocorrelation functions
of the eigenmodes.
%
For the single-bin (WL) scheme,
the optimal schedule from our method
is given by the inverse-time formula,
Eq. \eqref{eq:alpha_invt1}.
%
For a general multiple-bin scheme,
including the Gaussian one used in metadynamics,
the optimal schedule has no closed form,
but is given by an equation of motion
of a free particle with a position-dependent mass, Eq. \eqref{eq:q_opt}.
%
In this case,
the optimal schedule and error
can be sensitive to the simulation length.
%
The initial updating magnitude is always
half of the previous equilibrium value.
% the value used during equilibration.

We have also shown that
the single-bin scheme is optimal
in the long time limit
without a priori assumption
on the smoothness of the PMF.
%
More generally, with perfect filtering
of certain eigenmodes, we can construct a
class of bandpass updating schemes
that are also optimal for asymptotic convergence.
%
The optimal schedule of these bandpass schemes
is always given by the inverse-time formula.
%
They appeared to outperform similar Gaussian schemes
in our model study,
and they can be useful
for high-precision metadynamics simulations.
%
However, since the bandpass schemes
update the bias potential globally,
they can be more expensive than
the Gaussian schemes.
%
Besides, the performance gain
is relative small for a shorter simulation
with a local sampling process such as MD.

@@

We thanks
Dr. Y. Mei, Dr. J. Ma,
J. Drake, O. Nassar, Dr. C. Lai and Dr. S. Ou
for helpful discussions.

@@

As an example, we used the method to compute
the integrals, $\Gamma_k$'s, from an MD simulation,
and compared them to those from
Eq. \eqref{eq:Gamma_onestep}.
%the one-step sampling process.
%
As shown in Fig. \ref{fig:gamcmp},
the larger values of $\Gamma_k$'s
with smaller $k$'s
from the two sampling processes
were almost proportional,
suggesting similarity
of local sampling processes.

@@

%A practical problem of using homogeneous updating scheme
%is the following.
%
A homogeneous updating scheme
specified by an arbitrary kernel
is not necessarily stable,
i.e., some eigenvalues may be negative.
%
%
%\note{
  %For example,
  %the triple-bin updating scheme
  %is unstable if Eq. \eqref{eq:nn_stable} is violated.
  %
  Even
  the Gaussian updating scheme can be unstable
  if the updating kernel is truncated.
%}%
%
Here we show how to minimally modify
the updating kernel
to stabilize it.
%
\note{The relative code is \texttt{trimwindow()} in \texttt{invt.h}.
}

@@

\section{\label{sec:invt_schedule}
Inverse-time schedules}

>>>>

Here we shall call a simulation that targets
a nearly flat distribution\cite{
dayal2004, trebst2004, zhang2007, barducci2008, singh2011},
$p_i$,
a flat-distribution sampling (FDS) simulation.
%
Roughly speaking, there are two types of FDS methods.
%
The first type,
referred to as the equilibrium FDS, % (E-FDS) methods hereinafter,
uses a fixed bias potential, $V_i$,
(which is estimated beforehand)
during simulation.
%
Examples include the original
entropic or multicanonical sampling\cite{
berg1992, lee1993}
and simulated tempering\cite{
marinari1992, lyubartsev1992}
with $i$ being the index of energy
and that of temperature, respectively.
%
The estimated bias potential, $V_i$,
are often improved
by the histogram, $H_i$, along $i$ accumulated
during simulation.
%
Using $H_i$ as an estimate of $\pi_i$ in
Eq. \eqref{eq:pi_p_phi1}
allows us to correct $V_i$ as
%
\begin{equation}
V^\mathrm{corr}_i
=
V_i
+
\ln \frac{ H_i }
         { p_i }.
\label{eq:vcorr_equil}
\end{equation}
%
\note{This follows from
  $$
  \begin{aligned}
    H_i \approx \pi_i
    &\propto p^*_i \, e^{-V_i},
    \\
    p_i
    &\propto p^*_i \, e^{-V^\mathrm{corr}_i}.
  \end{aligned}
  $$
}
For an unknown system,
it is often difficult
to guess a sufficiently accurate
$V_i$ to start with.
%
Thus,
it can take many iterative rounds of simulations
to achieve the desired flat distribution.
%
\note{Further, a reasonably precise histogram
requires an accumulation period
longer than the autocorrelation time,
which can also be hard to estimate beforehand.}

@@

In our test system,
$i$ represents a one-dimensional
non-periodic variable,
and $n = 100$.
%
Both the target and intrinsic distributions
are assumed to be flat:
$p^*_i = p_i = 1/n$.
%
%We wish to see how
%the schedule, $\alpha(t)$,
%affects the fluctuating error
%of the bias potential
%at the end of the simulation.
%
Initially,
we set the bias potential to zero,
$V_i \equiv 0$,
and equilibrate the system for $10^7$ steps
under a constant updating magnitude,
$\alpha_0 = 10^{-4}$.
%
After this,
we reset the origin of the time, $t$,
and start the testing schedule
for $T = 10^8$ steps.
%
We repeat the above process $1000$ times
and average the results.
%specified by either
%Eq. \eqref{eq:q_opt}
%or
%Eq. \eqref{eq:alpha_invtlambda}.
Unless specified otherwise,
the above parameters were used.

@@

We performed adaptive FDS simulations
using the single-bin updating scheme
and the triple-bin updating scheme with $\mu_1 = 0.24$.
%
%For each value of $\lambda$,
%we ran $1000$ independent simulations
%of $10^8$ steps, and averaged the results.
%
As shown in Fig. \ref{fig:errinvt},
the errors from the simulations
agreed well those from the theory.

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=1.0\linewidth]{fig/errinvt.pdf}
  }
  \caption{
    \label{fig:errinvt}
    Error $E$ versus $\lambda$
    for the modified inverse-time schedule,
    Eq. \eqref{eq:alpha_invtlambda}.
    %
    The points are results from averaging over $1000$ independent runs;
    the curves are theoretical predictions.
    %
    \note{The figure was produced by \texttt{doc/fig/errinvt.gp}.
      The data were produced by the script \texttt{data/invtcscan.py},
      and saved under \texttt{data/invt}
      (see the \texttt{Makefile} there).
    }%
  }
\end{center}
\end{figure}

@@

To demonstrate the optimal schedule from Eq. \eqref{eq:q_opt},
we used the Gaussian (with $\sigma = 10$) updating scheme
as an example.
%
In Table \ref{tab:error_Gaussian},
we showed that the optimal schedule
produced smaller errors than
the inverse-time schedules.
%
However, the optimal schedules
decayed more slowly than
the inverse-time formula, Eq. \eqref{eq:alpha_invt1},
and depended on the sampling process,
as shown in Fig. \ref{fig:optacmp}.
%
The impeded decay of the optimal schedule
was necessary in order to
damp out the short-wavelength eigenmodes
(i.e., local noises),
as shown in Fig. \ref{fig:xerr}.
%
If, however, we could ignore these local eigenmodes
(by explicitly filtering out
them in the biased potential),
then we can modify the optimal schedule
from Eq. \eqref{eq:q_opt}
and limit the sum
to the first few eigenmodes.
%
The resulting schedule
would became similar to the inverse-time schedule,
Eq. \eqref{eq:alpha_invt1},
as shown in Fig. \ref{fig:optacmp}.
%
\note{A more thorough approach is
to use the bandpass updating scheme
to be discussed later.
}

@@

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=1.0\linewidth]{fig/optacmp.pdf}
  }
  \caption{
    \label{fig:optacmp}
    Optimal schedules from Eq. \eqref{eq:q_opt}
    for the Gaussian updating scheme
    with $\sigma = 10$.
    %
    The modified schedules, produced by
    limiting the sum in Eq. \eqref{eq:q_opt}
    to the first four terms,
    became similar to the inverse-time schedule,
    Eq. \eqref{eq:alpha_invt1}.
    %
    \note{This figure was produced by \texttt{doc/fig/optacmp.gp}.
      The data were computed by the program \texttt{prog/predict},
      see particularly the function \texttt{intq\_save()}
      in \texttt{prog/intq.h}.
      The data were saved under the directory \texttt{data/opta}.
    }%
  }
\end{center}
\end{figure}

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=1.0\linewidth]{fig/xerr.pdf}
  }
  \caption{
    \label{fig:xerr}
    Error components for the Gaussian updating scheme
    with $\sigma = 10$
    from the optimal schedule
    and the inverse-time schedule, Eq. \eqref{eq:alpha_invt1}.
    %
    The latter schedule failed to damp out the errors
    from the long-wavelength (local) eigenmodes
    with $K \ge 5$.
    %
    For longer simulations,
    the error components from the optimal schedule
    would roughly maintain the same shape,
    but ``intersect'' with the initial error curve
    at a lower level (not shown).
    %
    \note{This figure was produced by \texttt{doc/fig/xerr.gp}.
      The data were saved under the directory \texttt{data/xerr}.
    }%
  }
\end{center}
\end{figure}

To verify Eq. \eqref{eq:half_alpha0},
we computed the optimal schedules
for different values of $q(T)$,
and plotted the final error
against the initial updating magnitude, $\alpha(0)$.
%
As shown in Fig. \ref{fig:erriascan},
The minimal error was reached at
$\alpha(0) = \alpha/2 = 5 \times 10^{-5}$.

@@

In contrast,
the normalized errors of the bandpass updating schemes
were insensitive to the simulation length,
in agreement with Eq. \eqref{eq:error_sinc}.
%
Besides, the bandpass updating schemes
yielded smaller errors than
the corresponding Gaussian ones,
although the differences tended to
be small for shorter simulations
with the local sampling process.

@@

Each adaptive FDS method is associated with
an updating scheme,
or mathematically an updating matrix.
%whose elements $w_{ij}$ specifies the
%relative magnitude of updating the bias potential
%at bin $i$ upon a visit to bin $j$.
%
Our method computes the optimal schedule from
the eigenvalues of the updating matrix,
and the integrals of the autocorrelation functions
of the eigenmodes.
%
For the single-bin (WL) scheme,
the optimal schedule from our method
is given by the inverse-time formula,
Eq. \eqref{eq:alpha_invt1}.
%
For a general multiple-bin scheme,
including the Gaussian one used in metadynamics,
the optimal schedule permits no simple closed form,
but is given by an equation of motion
of a free particle with a position-dependent mass, Eq. \eqref{eq:q_opt}.
%
In this case,
the optimal schedule and error
can be sensitive to the simulation length.
%
The initial updating magnitude is always
half of the previous equilibrium value.
% the value used during equilibration.

We have also shown that
the single-bin scheme is optimal
in the long time limit
without a priori assumption
on the smoothness of the PMF.
%
More generally, with perfect filtering
of certain eigenmodes, we can construct a
class of bandpass updating schemes
that are also optimal for asymptotic convergence.
%
The optimal schedule of these bandpass schemes
is always given by the inverse-time formula.

Although the bandpass updating schemes
appeared to outperform similar Gaussian schemes
in our model study,
they are global in nature
and hence more expensive.
%
Besides, the performance gain
is relative small for a shorter simulation
with a local sampling process such as MD.
%
\note{
One way to improve the Gaussian updating schemes
is to use it with a hard cutoff of eigenmodes
(as we did for Fig. \ref{fig:optacmp}).
%
Then the optimal schedule
would be similar to the inverse-time one,
and the error, with the short-wavelength eigenmodes filtered,
would be close to that of the bandpass scheme.
}

@@

We thank Dr. J. Ma, Dr. Y. Mei,
J. Drake, O. Nassar, Dr. C. Lai and Dr. S. Ou
for helpful discussions.

@@

As an example, we used the method to compute
the integrals, $\Gamma_k$'s, from an MD simulation,
and compared them to those from
Eq. \eqref{eq:Gamma_onestep}.
%the one-step sampling process.
%
As shown in Fig. \ref{fig:gamcmp},
the larger values of $\Gamma_k$'s
with smaller $k$'s
from the two sampling processes
were almost proportional.
%
This suggests that
the optimal schedule
from simulations using local sampling processes
would be similar in general properties.

@@

%A practical problem of using homogeneous updating scheme
%is the following.
%
A homogeneous updating scheme
specified by an arbitrary kernel
is not necessarily stable,
i.e., some eigenvalues may be negative.
%
%
%\note{
  %For example,
  %the triple-bin updating scheme
  %is unstable if Eq. \eqref{eq:nn_stable} is violated.
  %
  Even
  the Gaussian updating scheme can be unstable
  in practice
  if the updating kernel is truncated.
%}%
%
Here we show how to minimally modify
the updating kernel
to stabilize it.
%
\note{The relative code is \texttt{trimwindow()} in \texttt{invt.h}.
}

@@

\section{\label{sec:invt_schedule}
Modified inverse-time schedules}
